Data Specialization for a Staged Lambda Calculus

### 2015-01-22 paper/splitting.tex
introduced a simple staged language

@@

Given a term $e$ in the source language \lang, where terms of
stage~\bbone\ and stage~\bbtwo\ may be interleaved, \emph{(stage)
  splitting} separates the stage~\bbone\ and stage~\bbtwo\
subcomputations into two terms,~$e_1$ and~$e_2$, in an unstaged target language \langmono\ 
%each performing the
%subcomputations corresponding to each stage.  More precisely, given a
%\lang\ term~$e$, splitting yields two terms $e_1$ and $e_2$ in a
%unstaged target language \langmono\
such that~$e_1$ performs all the
stage~\bbone\ subcomputations in~$e$ and~$e_2$~performs all of the
stage~\bbtwo\ subcomputations. Splitting preserves correctness in the sense that evaluating~$e$ in
the staged language produces the same result as evaluating~$e_1$ in the target language,
then further evaluating~$e_2$ in the target language with the result from~$e_1$.

In this section, we present a
splitting algorithm for the source language \lang, specifying
precisely the algorithm and its correctness conditions.  
We begin by defining a masking relation that serves to extract stage~\bbone\ and
stage~\bbtwo\ components of \lang\ partial values into \langmono\ terms that the outputs of splitting are designed to match. We then 
move on to the specification of the splitting algorithm itself.

>>>>

%!TEX root = paper.tex

@@

The primary strength of \lang\ is the expressive power afforded by
its ability to syntactically inteleave work intended for stage \bbone\
and work intended for stage \bbtwo, as well as it ability to form
mixed-stage abstractions.

But one could imagine a simpler multistage language, \langS,
wherein the stages are syntactically separated {\em in the source}.
That is, there is only one multistage construct, and it has the form:
\[
\dothen {e_1} x {e_2}
\]
where $e_1$ and $e_2$ are entirely monostage terms that
encode the stage \bbone\ work and stage \bbtwo\ work, respectively,
of the whole program.
The type system for this language has only a single multistage type,
$A \& B$ which means the program produces a value of type $A$ 
as the output of stage \bbone\ and a value of type $B$ as the
output of stage \bbtwo.  
There is also only one typing rule:
\[
\infer {\dothen {e_1} x {e_2} : A \& B} {\types[\emptyC]{e_1}{A\times \tau} & \types[x : \tau]{e_2}{B}}
\]
Phase \bbone\ evaluation is given by
\[
\infer {\dothen {e_1} x {e_2} \Downarrow_\bbone^{S} [v_1; [u/x]e_2] } {\reduce{e_1}{(v_1,u)}}
\]
and phase \bbtwo\ evaluation is given by
\[
\infer {[v_1; e_2] \Downarrow_\bbtwo^{S} [v_1; v_2] } {\reduce{e_2}{v_2}}
\]

[Need to talk about the other flavor in \langS, which produces no first-stage output.]

% Given a term $e$ in the source language \lang, where terms of
% stage~\bbone\ and stage~\bbtwo\ may be interleaved, \emph{(stage)
%   splitting} separates the stage~\bbone\ and stage~\bbtwo\
% subcomputations into two terms,~$e_1$ and~$e_2$, in an unstaged target language \langmono\ 
% %each performing the
% %subcomputations corresponding to each stage.  More precisely, given a
% %\lang\ term~$e$, splitting yields two terms $e_1$ and $e_2$ in a
% %unstaged target language \langmono\
% such that~$e_1$ performs all the
% stage~\bbone\ subcomputations in~$e$ and~$e_2$~performs all of the
% stage~\bbtwo\ subcomputations. Splitting preserves correctness in the sense that evaluating~$e$ in
% the staged language produces the same result as evaluating~$e_1$ in the target language,
% then further evaluating~$e_2$ in the target language with the result from~$e_1$.

The goal of stage splitting is to translate terms of \lang\ into equivalent terms of \langS.
[Specifically, there are versions for each flavor.]

We begin by defining {\em masking} operations that relate the full and partial values of \lang\ and \langS.
With our end goal defined, we move on to the specification of the term splitting algorithm itself.
